---
title: "GO Analysis"
author: "Ha Tran"
date: "22/08/2021"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  fig.align = "center"
)
```

# Data Setup

```{r load libraries}
# working with data
library(dplyr)
library(magrittr)
library(readr)
library(tibble)
library(reshape2)
library(tidyverse)

# Visualisation:
library(kableExtra)
library(ggplot2)
library(grid)

# Custom ggplot
library(gridExtra)
library(ggbiplot)
library(ggrepel)


# Bioconductor packages:
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)

theme_set(theme_minimal())
pub <- readRDS(here::here("0_data/RDS_objects/pub.rds"))
```


### Import DGElist Data

DGElist object containing the raw feature count, sample metadata, and gene metadata, created in the Set Up stage. 

```{r importData}
# load DGElist previously created in the set up
dge <- readRDS(here::here("0_data/RDS_objects/dge.rds"))
fc <- readRDS(here::here("0_data/RDS_objects/fc.rds"))
lfc <- readRDS(here::here("0_data/RDS_objects/lfc.rds"))
lmTreat <- readRDS(here::here("0_data/RDS_objects/lmTreat.rds"))
lmTreat_sig <- readRDS(here::here("0_data/RDS_objects/lmTreat_sig.rds"))
```


# GO Analysis {.tabset}
`goSummaries` is a package created by Dr Stephen Pederson for filtering GO terms based on ontology level.

```{r}
# circumvent rerunning of lengthy analysis.
enrichGO <- readRDS(here::here("0_data/RDS_objects/enrichGO.rds"))
enrichGO_sig <- readRDS(here::here("0_data/RDS_objects/enrichGO_sig.rds"))
```

```{r enrichGO, eval=FALSE}
# download go summaries and set the minimum ontology level
goSummaries <- url("https://uofabioinformaticshub.github.io/summaries2GO/data/goSummaries.RDS") %>%
  readRDS()
minPath <- 4

enrichGO=list()
enrichGO_sig <- list()
for (i in 1:length(fc)) {
  x <- fc[i] %>% as.character()
  # find enriched GO terms
  enrichGO[[x]] <- clusterProfiler::enrichGO(
    gene = lmTreat_sig[[x]]$entrezid,
    OrgDb = org.Mm.eg.db,
    keyType = "ENTREZID",
    ont = "ALL",
    pAdjustMethod = "fdr",
    pvalueCutoff = 0.05
  )

  # bind to goSummaries to elminate go terms with ontology levels 1 and 2.
  enrichGO_sig[[x]] <- enrichGO[[x]] %>% 
    clusterProfiler::setReadable(OrgDb = org.Mm.eg.db, keyType = "auto")

  enrichGO_sig[[x]] <- enrichGO_sig[[x]] %>%
    as.data.frame() %>%
    rownames_to_column("id") %>%
    left_join(goSummaries) %>%
    dplyr::filter(shortest_path >= minPath) %>%
    column_to_rownames("id")

  # adjust go results, separate compound column, add FDR column, adjust the GeneRatio column
  enrichGO_sig[[x]] <- enrichGO_sig[[x]] %>%
    separate(col = BgRatio, sep = "/", into = c("Total", "Universe")) %>%
    dplyr::mutate(
      logFDR = -log(p.adjust, 10),
      GeneRatio = Count / as.numeric(Total)) %>% 
    dplyr::select(c("Description", "ontology", "GeneRatio", "pvalue", "p.adjust", "logFDR", "qvalue", "geneID"))
  
  # at the beginnning of a word (after 35 characters), add a newline. shorten the y axis for dot plot 
  enrichGO_sig[[x]]$Description <- sub(pattern = "(.{1,35})(?:$| )", 
                                       replacement = "\\1\n", 
                                       x = enrichGO_sig[[x]]$Description)
  
  # remove the additional newline at the end of the string
  enrichGO_sig[[x]]$Description <- sub(pattern = "\n$", 
                                       replacement = "", 
                                       x = enrichGO_sig[[x]]$Description)
}
```


## FC=`r fc[1]`

```{r}
# display the top 30 most sig
enrichGO_sig[[1]] %>% 
  dplyr::mutate_if(is.numeric, funs(as.character(signif(.,3)))) %>% 
  kable(caption = "Significantly enriched GO terms") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "600px")
```

### Visualisation {.tabset}
```{r visualisationSetup}
bp_dot=list()
mf_dot=list()
cc_dot=list()
upset=list()
for (i in 1:length(fc)) {
  x <- fc[i] %>% as.character()
  
  # extract the enriched GO terms from each ontology
  bp <- enrichGO_sig[[x]] %>% dplyr::filter(ontology == "BP") 
  mf <- enrichGO_sig[[x]] %>% dplyr::filter(ontology == "MF")
  cc <- enrichGO_sig[[x]] %>% dplyr::filter(ontology == "CC")

  # bp dot plot, save
  bp_dot[[x]] <- ggplot(bp[1:15, ]) +
    geom_point(aes(x = GeneRatio, y = reorder(Description, GeneRatio), colour = logFDR, size = Count)) +
    scale_color_gradient(low = "dodgerblue3", high = "firebrick3", limits = c(0, NA)) +
    scale_size(range = c(1.5,5)) +
    ggtitle("Biological Process") +
    ylab(label = "") +
    xlab(label = "Gene Ratio") +
    labs(color = expression("-log"[10] * "FDR"), size = "Gene Counts")
  ggsave(filename = paste0("bp_dot_", fc[i], ".svg"), plot = bp_dot[[x]] + pub, path = here::here("2_plots/go/"), 
         width = 250, height = 166, units = "mm")

  # mf dot plot, save
  mf_dot[[x]] <- ggplot(mf[1:15, ]) +
    geom_point(aes(x = GeneRatio, y = reorder(Description, GeneRatio), colour = logFDR, size = Count)) +
    scale_color_gradient(low = "dodgerblue3", high = "firebrick3", limits = c(0, NA)) +
    scale_size(range = c(1.5,5)) +
    ggtitle("Molecular Function") +
    ylab(label = "") +
    xlab(label = "Gene Ratio") +
    labs(color = expression("-log"[10] * "FDR"), size = "Gene Counts")
  ggsave(filename = paste0("mf_dot_", fc[i], ".svg"), plot = mf_dot[[x]] + pub, path = here::here("2_plots/go/"), 
         width = 250, height = 166, units = "mm")

  # cc dot plot, save
  cc_dot[[x]] <- ggplot(cc[1:15, ]) +
    geom_point(aes(x = GeneRatio, y = reorder(Description, GeneRatio), colour = logFDR, size = Count)) +
    scale_color_gradient(low = "dodgerblue3", high = "firebrick3", limits = c(0, NA)) +
    scale_size(range = c(1.5,5)) +
    ggtitle("Cellular Components") +
    ylab(label = "") +
    xlab(label = "Gene Ratio") +
    labs(color = expression("-log"[10] * "FDR"), size = "Gene Counts")
  ggsave(filename = paste0("cc_dot_", fc[i], ".svg"), plot = cc_dot[[x]] + pub, path = here::here("2_plots/go/"), 
         width = 250, height = 166, units = "mm")
  
  upset[[x]] <- upsetplot(x = enrichGO[[x]], 10)
  ggsave(filename = paste0("upset_", fc[i], ".svg"), plot = upset[[x]], path = here::here("2_plots/go/"), width = 250, height = 166, units = "mm")
}
```

#### Biological Process

```{r}
bp_dot[[1]] 
```

#### Molecular Function

```{r}
mf_dot[[1]]
```

#### Cellular Components

```{r}
cc_dot[[1]]
```

#### Upset

```{r}
upset[[1]]
```

## FC=`r fc[2]`

```{r}
# display the top 30 most sig
enrichGO_sig[[2]] %>% 
  dplyr::mutate_if(is.numeric, funs(as.character(signif(.,3)))) %>% 
  kable(caption = "Significantly enriched GO terms") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "600px")
```

### Visualisation {.tabset}

#### Biological Process

```{r}
bp_dot[[2]]
```

#### Molecular Function

```{r}
mf_dot[[2]]
```

#### Cellular Components

```{r}
cc_dot[[2]]
```

#### Upset

```{r}
upset[[2]]
```

## FC=`r fc[3]`

```{r}
# display the top 30 most sig
enrichGO_sig[[3]] %>% 
  dplyr::mutate_if(is.numeric, funs(as.character(signif(.,3)))) %>% 
  kable(caption = "Significantly enriched GO terms") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "600px")
```

### Visualisation {.tabset}

#### Biological Process

```{r}
bp_dot[[3]]
```

#### Molecular Function

```{r}
mf_dot[[3]]
```

#### Cellular Components

```{r}
cc_dot[[3]]
```

#### Upset

```{r}
upset[[3]]
```


# Export Data
```{r exportData}
# save to excel
writexl::write_xlsx(x = enrichGO_sig, here::here("3_output/enrichGO_sig.xlsx"))
saveRDS(object = enrichGO_sig,file = here::here("0_data/RDS_objects/enrichGO_sig.rds"))
saveRDS(object = enrichGO,file = here::here("0_data/RDS_objects/enrichGO.rds"))

```
