---
title: "Setting up for DGE analysis (Robertson Lab)"
author: "Ha Tran"
date: "02/09/2021"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 11)
```

# Data Setup

Prior to this analysis, reads were:
1. Trimmed using `AdapterRemoval`
2. Aligned to **GRCm38/mm10** using `STAR`
3. Reads quantification performed with `featureCounts`

**Transcript QC, alignment, and quantification were performed by Dr Jimmy Breen**

```{r loadLibraries}
# working with data
library(dplyr)
library(magrittr)
library(readr)
library(tibble)
library(reshape2)
library(tidyverse)
library(bookdown)

# Visualisation:
library(kableExtra)
library(ggbiplot)
library(ggrepel)
library(grid)
library(cowplot)
library(corrplot)

# Set ggplot theme
theme_set(theme_minimal())

pub <- theme_update(
  plot.title = element_text(color = "gray20", size = 7, angle = 0, hjust = 0.5, vjust = .5, face = "bold"),
  plot.subtitle = element_text(color = "gray20", size = 6, angle = 0, hjust = 0, vjust = .5, face = "plain"),
  legend.title = element_text(color = "gray20", size = 5, angle = 0, hjust = 0.5, vjust = .5, face = "plain"),
  legend.text = element_text(color = "gray20", size = 5, angle = 0, hjust = 0, vjust = .5, face = "plain"),
  axis.text.x = element_text(color = "gray20", size = 5, angle = 0, hjust = .5, vjust = 0, face = "plain"),
  axis.title.x = element_text(color = "gray20", size = 6, angle = 0, hjust = .5, vjust = 0, face = "plain"),
  axis.text.y = element_text(color = "gray20", size = 5, angle = 0, hjust = 1, vjust = 0.5, face = "plain"),
  axis.title.y = element_text(color = "gray20", size = 6, angle = 90, hjust = .5, vjust = .5, face = "plain"))

# Bioconductor packages:
library(AnnotationHub)
library(edgeR)
library(limma)
library(Glimma)
```

## Import Raw Count Data

Due to the unusual library size, control 3 was removed from the analysis

```{r importRawCountData}
# import the mergedOnly dataset, provided by Dr Jimmy Breen on the 24/09/21
rawCount <- read_tsv(here::here("0_data/raw_data/allSamples_mergedOnly.featureCounts.txt"),
                     col_names = TRUE,
                     comment = "#") %>%
  dplyr::rename(CONT1 = "../2_Hisat2_merged/CONT1_ATGTCA_merged.sorted.nodup.bam",
                CONT2 = "../2_Hisat2_merged/CONT2_CGATGT_merged.sorted.nodup.bam",
                CONT4 = "../2_Hisat2_merged/CONT4_ACTTGA_merged.sorted.nodup.bam",
                INT1 = "../2_Hisat2_merged/INT1_GTCCGC_merged.sorted.nodup.bam",
                INT2 = "../2_Hisat2_merged/INT2_ACAGTG_merged.sorted.nodup.bam",
                INT3 = "../2_Hisat2_merged/INT3_GATCAG_merged.sorted.nodup.bam",
                INT4 = "../2_Hisat2_merged/INT4_CTTGTA_merged.sorted.nodup.bam",
                SVX1 = "../2_Hisat2_merged/SVX1_GTTTCG_merged.sorted.nodup.bam",
                SVX2 = "../2_Hisat2_merged/SVX2_TAGCTT_merged.sorted.nodup.bam",
                SVX3 = "../2_Hisat2_merged/SVX3_ATCACG_merged.sorted.nodup.bam",
                SVX4 = "../2_Hisat2_merged/SVX4_GCCAAT_merged.sorted.nodup.bam",
                SVX_VAS1 = "../2_Hisat2_merged/SVX_VAS1_AGTCAA_merged.sorted.nodup.bam",
                SVX_VAS2 = "../2_Hisat2_merged/SVX_VAS2_AGTTCC_merged.sorted.nodup.bam",
                SVX_VAS3 = "../2_Hisat2_merged/SVX_VAS3_TGACCA_merged.sorted.nodup.bam",
                SVX_VAS4 = "../2_Hisat2_merged/SVX_VAS4_GGCTAC_merged.sorted.nodup.bam",
                VAS1 = "../2_Hisat2_merged/VAS1_CAGATC_merged.sorted.nodup.bam",
                VAS2 = "../2_Hisat2_merged/VAS2_GTGAAA_merged.sorted.nodup.bam",
                VAS3 = "../2_Hisat2_merged/VAS3_GTGGCC_merged.sorted.nodup.bam",
                VAS4 = "../2_Hisat2_merged/VAS4_CCGTCC_merged.sorted.nodup.bam",) %>% 
  column_to_rownames("Geneid") %>%
  as.data.frame() 
  rownames(rawCount) <- gsub("\\..+$", "", rownames(rawCount))
  
# Removing the non-numerical metadata column. SVX_VAS1 may also be an outlier, it is number 18 (BTW)
rawCount<- rawCount[, c(6,7,9:25)]
```

## Import Metadata

There are generally two metadata required for DGE analysis.

1.  metadata about *each sample*

2.  metadata about *each gene*

### Sample Metadata

The sample metadata can be extracted from the `logCPM` column names. These data include `sample_id`, `sample_group`, `sample_type`.

The sample metadata will be manually generated and stored in the `/0_data/raw_data/` directory

```{r sampleMetadata}
samples <- read_tsv(here::here("0_data/raw_data/samples.tsv"),
                    col_names = TRUE) %>% 
  column_to_rownames("1")
```

### Gene Metadata

Gene annotation is useful for the DGE analysis as it will provide useful information about the genes. The annotated genes of *Mus musculus* can be pulled down by using Annotation Hub.

Annotation Hub also has a web service that can be assessed through the display function. Pulling down the gene annotation can take a long time, so after the initial run, the annotated genes is saved to a `genes.rds` file. To save time, if `genes.rds` is already present, don't run the code chunk.

```{r AnnotationHub, eval=FALSE}
ah <- AnnotationHub()
ah %>% 
  subset(grepl("musculus", species)) %>%
  subset(rdataclass == "EnsDb")

#viewing web service for annotation hub 
#d <- display(ah)

# Annotation hub html site was used to identify 'code' for the latest mouse genome from Ensembl
ensDb <- ah[["AH95775"]]

genes <- genes(ensDb) %>% 
  as.data.frame()

#the annotated genes are saved into a RDS object to save computational time in subsequent run of the setUp.Rmd
genes %>% saveRDS(here::here("0_data/RDS_objects/gene_metadata.rds"))
```

Using the annotated gene list through AnnotationHub(), load into object called `geneMetadata`. Filter out all genes that are present in the rawCount and display the number of unique gene_biotypes present in the rawCount and geneMetadata

```{r geneAnnotation, paged.print=5}
geneMetadata <- read_rds(here::here("0_data/RDS_objects/gene_metadata.rds"))

#prepare the gene data frame to contain the genes listed in the rownames of 'rawCount' data
geneMetadata <- data.frame(gene = rownames(rawCount)) %>% 
  left_join(geneMetadata %>% as.data.frame, 
            by = c("gene"="gene_id")) %>% 
  dplyr::distinct(gene, .keep_all=TRUE)  

rownames(geneMetadata) <- geneMetadata$gene

#Using the table function, the details of the genes present in the rawCount data can be summaried.
genes <- geneMetadata$gene_biotype %>% table %>% as.data.frame()

colnames(genes) <- c("Gene Biotype", "Frequency")
  
kable(genes) %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height = "600px")
```

## Create DGEList object

Digital Gene Expression List (DGElist) is a R object class often used for differential gene expression analysis as it simplifies plotting, and interaction with data and metadata.

The DGEList object holds the three dataset that have imported/created, including `rawCount` data and `sampleMetadata` and `geneMetadata` metadata.

To further save time and memory, genes that were not expressed across all samples (i.e., `0` count across all columns) are all removed

```{r createDGElist}
#Create DGElist with rawCOunt and gene data. Remove all genes with 0 expression in all treatment groups
dge <- DGEList(counts = rawCount, 
               samples = samples,
               genes = geneMetadata,
               remove.zeros = TRUE
               ) 
```

# Pre-processing and QC

Pre-processing steps increased the power of the downstream DGE analysis by eliminating majority of unwanted variance that could obscure the true variance caused by the differences in sample conditions. There are several standard steps that are commonly followed to pre-process and QC raw read counts, including:

1.  Checking Library Size

2.  Removal of Undetectable Genes

3.  Normalisation

4.  QC through MDS/PCA

## Checking Library Size

A simple pre-processing/QC step is checking the quality of library size (total number of mapped and quantified reads) for each treatment. This enable identification of potentially mis-labelled or outlying samples. This is often visualised through `ggplot`.

```{r libraryChecks, fig.cap = "Sample library size. Dash line represent average library size"}
libSize <- dge$samples %>% 
  
  #plot the sample with the lib.size in x and sample_group in y, colour fill for each sample_group
  ggplot(aes(
    x = lib.size, 
    y = rownames(dge$samples), 
    fill = dge$samples$group)
    ) +
  geom_col() +
  
  #draw a vertical line for the mean lib.size
  geom_vline(
    aes (xintercept = lib.size), 
    data = . %>% summarise_at(vars(lib.size), mean),
    linetype = 2
  ) +
  
  #labelling splot
  labs(
    title = "Sample Library Size",
    x = "Library Size",
    y = "Samples",
    fill = "Sample Groups"
  ) +
  
  #PUBLISHING
  theme(legend.position = "none")

libSize

#save the plot to .svg
ggsave(here::here("2_plots/qc/library_size.svg"),
       plot = libSize + pub,
       #PUBLISHING
       width = 250,
       height = 166,
       units = "mm")
```


## Removal of Low-Expressed Genes

Filtering out low-expressed genes is a standard pre-processing step in DGE analysis as it can significantly increase the power to differentiate differentially expressed genes by eliminating the variance caused by genes that are lowly expressed in all samples.

The threshold of removal is arbitrary and is often determined after visualisation of the count distribution. The count distribution can be illustrated in a density plot through `ggplot`. A common metric used to display the count distribution is *log Counts per Million* (logCPM)

```{r beforeFiltering, results='hide'}
beforeFiltering <- dge %>% 
  
  #transform the raw count to logCPM 
  edgeR::cpm(log = TRUE) %>% 

  #melting (reorganising) the transformed logCPM data with respect to the id variable (i.e., the row and column names). Very similar to pivot_long function
  melt %>%

  #retain all rows where the logCPM (value) is finite. All of them in this case are finite
  dplyr::filter(is.finite(value)) %>%

  #plot the long formate logCPM counts in a density plot with x the logCPM and colour by the sample_id
  ggplot(aes(
    x = value,
    colour = Var2
  )) +
  geom_density() +

  # remove the legend
  guides(colour = FALSE) +

  #add figure tittle and subtitle and labels
  ggtitle("Before Filtering Low-expressed Genes", subtitle = paste0(nrow(dge), " genes"))+
  labs(
    x = "logCPM",
    y = "Density",
    colour = "Sample Groups"
  )

#save plot
ggsave("counts_before_filtering.svg",
       plot = beforeFiltering + pub,
       width = 250,
       height = 166,
       units = "mm",
       path = here::here("2_plots/qc/"))

```

Ideally, the filtering the low-expressed genes should remove the large peak with `logCPM < 0`, i.e., remove any genes which have less than one count per million.

A common guideline is to keep all genes that have > 1-2 cpm in the smallest group on a treatment. In this case, the smallest group is 3 as each treatment condition had three replicates. However, due to the high variance of some groups, the filtering is increased to keep genes that are are **more than 3 CPM in at least 3 samples**. Mathematically this would be identifying genes (rows) with CPM `> 3`; and identifying total row sum that is `>= 3`.

```{r afterFiltering, fig.cap="Before and after removal of lowly expressed genes"}
#the genes kept have >2 CPM for at least 3 samples
keptGenes <- (rowSums(cpm(dge) > 3) >= 3) 
afterFiltering <- dge %>% 
  
  #transform the raw count to logCPM 
  edgeR::cpm(log = TRUE) %>% 
  
  #for var1 (gene names) extract only the keptGenes and discard all other genes in the logCPM data
  magrittr::extract(keptGenes,) %>% 

  #melting (reorganising) the transformed logCPM data with respect to the id variable (i.e., the row and column names). Very similar to pivot_long function
  melt %>% 
  
  #retain all rows where the logCPM (value) is finite. All of them in this case are finite
  dplyr::filter(is.finite(value)) %>% 

  #ggplot
  ggplot(aes(
    x = value,
    colour = Var2
  )) +
  geom_density() +
  
  #remove the legend
  guides(colour = FALSE) +
  
  #add figure tittle and subtitle and labels. since keptGenes is a logic element, the second element represents the number of genes that were kept after the filtering
  ggtitle("After Filtering Low-expressed Genes", subtitle = paste0(table(keptGenes)[[2]], " genes")) +
  labs(
    x = "logCPM",
    y = "Density",
    colour = "Sample Groups"
  )

#save plot
ggsave("counts_after_filtering_3_3.svg",
       plot = afterFiltering + pub,
       width = 250,
       height = 166,
       units = "mm",
       path = here::here("2_plots/qc/"))

#display plot
# afterFiltering

#display plot side by side
cowplot::plot_grid(beforeFiltering + pub, afterFiltering + pub)
ggsave(filename = "counts_before_after_filtering_3_3.svg",
        path = here::here("2_plots/qc/"),

       # PUBLISHING
       width = 320,
       height = 180,
       units = "mm")
```

Following the filtering of low-expressed genes **\< 3 CPM in at least 3 samples**, out of the total `r length(keptGenes)` genes left after the removal of genes with no expression, `r table(keptGenes)[[1]]` genes were removed, leaving only `r table(keptGenes)[[2]]` genes remaining for the downstream analysis

### Subset the DGElist object

After filtering the low-expressed genes, the DGElist object is updated to eliminate the low-expressed genes from future analysis

```{r subsetDGE}
#extract genes from keptGenes and recalculate the lib size
dge <- dge[keptGenes,,keep.lib.sizes = FALSE]
```


## Normalisation

Using the TMM (trimmed mean of M value) method of normalisation through the `edgeR` package. The TMM approach creates a scaling factor as an offset to be supplied to Negative Binomial model. The `ca;cNormFactors` function calculate the normalisation and return the adjusted `norm.factor` to the `dge$samples` element.

```{r afterNormalisation}
#after normalisation
dge <- edgeR::calcNormFactors(object = dge, 
                              method = "TMM")
knitr::kable(dge$samples, caption = "Normalised samples") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  scroll_box(height = "600px")
```

### Visualisation of TMM Normalisation

The following visualisation of the TMM normalisation is plotted using the mean-difference (MD) plot. The MD plot visualise the library size-adjusted logFC between two samples (the difference) against the log-expression across all samples (the mean). In this instance, `sample 1` is used to compare against an artificial library construct from the average of all the other samples

```{r visualisingNormalisation, fig.cap="MA plot of TMM normalisation for control 1"}
limma::plotMD(cpm(dge, log = TRUE), column=1)
abline(h=0, col="red", lty=2, lwd=2)
```

Ideally, the bulk of gene expression following the TMM normalisation should be centred around `expression log-ratio` of 0, which indicates that library size bias between samples have been successfully removed. This should be repeated with all the samples in the dge object.

## Quality Control

### Pinciple Component Analysis (PCA)

```{r pca, fig.cap="PCA plot of all samples."}
samples <- dge$samples %>% rownames_to_column("sampleName")
samples$rep <- samples$rep %>% as.factor()


# Perform PCA analysis:
pca_analysis <- prcomp(t(cpm(dge, log = TRUE)))

summary(pca_analysis)$importance %>% as.data.frame() 

group.colours <- c(CONT = "#F8766D", INT = "#A3A500")

# Create the plot
a <- pca_analysis$x %>% 
  as.data.frame() %>% 
  rownames_to_column("sampleName") %>% 
  left_join(samples) %>% 
  as_tibble() %>% 
  ggplot(aes(x = PC1, y = PC2, colour = group, shape = rep)) +
  geom_point(size=3, alpha=0.5) +
  scale_shape_manual(values = c(15:18)) +
  labs(
    x = paste0("PC1 (", percent(summary(pca_analysis)$importance["Proportion of Variance","PC1"]),")"),
    y = paste0("PC2 (", percent(summary(pca_analysis)$importance["Proportion of Variance","PC2"]),")"),
    colour = "Groups",
    shape = "Replicates"
  )

b <- pca_analysis$x %>% 
  as.data.frame() %>% 
  rownames_to_column("sampleName") %>% 
  left_join(samples) %>% 
  as_tibble() %>% 
  ggplot(aes(x = PC2, y = PC3, colour = group, shape = rep)) +
  geom_point(size=3, alpha=0.5) +
  scale_shape_manual(values = c(15:18)) +
  labs(
    x = paste0("PC2 (", percent(summary(pca_analysis)$importance["Proportion of Variance","PC2"]),")"),
    y = paste0("PC3 (", percent(summary(pca_analysis)$importance["Proportion of Variance","PC3"]),")"),
    colour = "Groups",
    shape = "Replicates"
  )

c <- pca_analysis$x %>% 
  as.data.frame() %>% 
  rownames_to_column("sampleName") %>% 
  left_join(samples) %>%
  as_tibble() %>%
  dplyr::slice(1:7) %>% 
  ggplot(aes(x = PC1, y = PC2, colour = group, shape = rep)) +
  geom_point(size=3, alpha=0.8) +
  scale_color_manual(values = group.colours)+
  scale_shape_manual(values = c(15:18)) +
  labs(
    x = paste0("PC1 (", percent(summary(pca_analysis)$importance["Proportion of Variance","PC1"]),")"),
    y = paste0("PC2 (", percent(summary(pca_analysis)$importance["Proportion of Variance","PC2"]),")"),
    colour = "Groups",
    shape = "Replicates"
  )

c

ggsave("PCA_IntvsCont.svg",
       plot = c + pub,
       path = here::here("2_plots/qc/"),
       width = 150,
       height = 100,
       units = "mm")

# pca_plot_2 <- plot_grid(
#     plot_grid(
#         a + theme(legend.position = "none"), 
#         b + theme(legend.position = "none"),
#         c + theme(legend.position = "none"),
#         nrow = 1
#     ),
#     get_legend(a + theme(legend.position = "bottom")),
#     nrow = 2, 
#     rel_heights = c(4,1)
# )
# 
# 
# pca_plot_2


# ggsave("PCA_plot.svg",
#        plot = pca_plot_2,
#        path = here::here("2_plots/qc/"),
#        width = 188,
#        height = 100,
#        units = "mm"
#        )
```

### Correlation plot

```{r corr_plot, fig.cap="Correlation between first three principle components and measured variables"}
corr_plot <- pca_analysis$x %>%
  as.data.frame() %>% 
  rownames_to_column("sampleName") %>% 
  left_join(samples) %>% 
  as_tibble() %>% 
  dplyr::select(
    PC1, 
    PC2, 
    PC3,
    Groups=group,
    Mated,
    "Library size"=lib.size,
    "Normalisation Factor"=norm.factors
  ) %>% 
  mutate(Groups = as.numeric(as.factor(Groups))) %>% 
  cor(method = "spearman") %>% 
  corrplot(
    type = "lower",
    diag = FALSE,
    addCoef.col = 1, addCoefasPercent = TRUE
  )
```

## Save DGElist object

```{r savingDGElist}
# Save DGElist object into the data/R directory 
saveRDS(object = dge, file = here::here("0_data/RDS_objects/dge.rds"))
saveRDS(object = pub, file = here::here("0_data/RDS_objects/pub.rds"))

# saveRDS(object = gg_publish, file = here::here("0_data/RDS_objects/gg_publish.rds"))
```
